Evaluating the Capabilities of the Lattice Boltzmann Method for non-Newtonian and Free-surface Flow towards Applications in Wellbore Cementing

### 2016-09-03 src/stability-paper.tex
added better conclusions at the end of the abstract and a simple analysis of gradients at the end of the poiseulle flow section

The Lattice Boltzmann method (LBM) is a numerical method based on computational statistical mechanics that is well-suited for approximating complex flow behaviors such as non-Newtonian, free surface, and multiphase multicomponent flow.
LBM is typically applied to simulate flow through a series of time steps, each consisting of streaming particle distributions to neighboring nodes and collisions of particle distributions at each node through a collision operator.
The collision operator is of interest because it, along with the equilibrium distribution function, determine the physics that are simulated (e.g. constitutive laws, interfacial dynamics, etc.) and it has implications on numerical stability and computational efficiency.
This work examines various collision operators and methods for stability enhancement for their suitability for simulating non-Newtonian fluid flows in terms of accuracy, numerical stability and computational efficiency.
The investigation was carried out as a numerical study looking for qualitative, yet practical, results, including testing the BGK and MRT collision operators, with and without entropic filtering, as applied to Bingham plastics and power-law fluids.
Two different benchmark problems were chosen for the flows: Poiseuille flow, and lid-driven square cavity flow.
The test results are followed by recommendations for choice collision scheme given priority and problem type.

@@

>>>>

The Lattice Boltzmann method (LBM) is a numerical method based on computational statistical mechanics that is well-suited for approximating complex flow behaviors such as non-Newtonian, free surface, and multiphase multicomponent flow.
LBM is typically applied to simulate flow through a series of time steps, each consisting of streaming particle distributions to neighboring nodes and collisions of particle distributions at each node through a collision operator.
The collision operator is of interest because it, along with the equilibrium distribution function, determine the physics that are simulated (e.g. constitutive laws, interfacial dynamics, etc.) and it has implications on numerical stability and computational efficiency.
This work examines various collision operators and methods for stability enhancement for their suitability for simulating non-Newtonian fluid flows in terms of accuracy, numerical stability and computational efficiency.
The investigation was carried out as a numerical study looking for qualitative, yet practical, results, including testing the BGK and MRT collision operators, with and without entropic filtering, as applied to Bingham plastics and power-law fluids.
Two different benchmark problems were chosen for the flows: Poiseuille flow, and lid-driven square cavity flow.
The results of the numerical study showed that the MRT collision operator can have an advantage in terms of stability and accuracy for a variety of non-Newtonian flow behavior, but at an increased computing cost that was, in some cases, as much as five times greater.
It was also shown that artificial dissipation can be an effective technique for enhancing numerical stability of non-Newtonian, lid-driven cavity flow, and is particularly effective for efficiently and accurately simulating the flow of shear-thinning fluids.

@@

Before closing the discussion on the Poiseuille flow numerical study, it is worthwhile to consider what precisely is common between the higher-yield stress fluids and the shear-thinning fluids, as they were the cases that exhibited numerical instabilities and were simulated with the least amount of accuracy for some of the collision schemes considered.
One such commonality can be seen when considering the effect of the yield stress, $\tau_y$, and the effect of the flow behavior index, $n$, on the gradient of the theoretical flow velocity.
Taking the partial with respect to $y$ of the analytical solution for the flow velocity of a Bingham plastic, Poiseuille flow (Equation \eqref{eq:analytical-bingham}) results in:
\begin{equation} \label{eq:gradu-bingham}
  \frac{\partial u_x}{\partial y} = \begin{cases}
    0, & 0 \leq |y| \leq y_{\tau}, \\
    \frac{1}{\mu_p} \left( \pgrad y + \tau_y \frac{y}{|y|} \right), & y_{\tau} < |y| \leq h,
  \end{cases}
\end{equation}
and taking the partial with respect to $y$ of the analytical solution for the flow velocity of a power-law fluid, Poiseuille flow (Equation \eqref{eq:analytical-power-law}) results in:
\begin{equation} \label{eq:gradu-power-law}
  \frac{\partial u_x}{\partial y} = \left( -\frac{1}{k} \pgrad \right)^{1/n} \left( -y |y|^{1/n - 1} \right).
\end{equation}
It can be seen from \eqref{eq:gradu-bingham} and \eqref{eq:gradu-power-law} that $\left| \frac{\partial u_x}{\partial y} \right|$ increases with the yield stress for Bingham plastics, and increases with $\frac{1}{n}$ for power-law fluids, which suggests that one of the challenges associated with the accuracy and stability of simulating higher yield-stress fluids and more extreme shear-thinning fluids (e.g. $n \le 0.5$) in Poiseuille-type flow is a result of having a larger magnitude of the gradient of macroscopic flow.
This conclusion is consistent with the discussion in \Fref{sec:bgk} with regards to one of the drawbacks of the BGK collision operator, namely that when there is a sharp spatial gradient in either $\rho$ or $\mvel$ that it can lead to nonphysical oscillations that can degrade the accuracy and stability of the solution.
However, it does not explain why the MRT collision operator was unable to accurately simulate the case of a power-law fluid, Poiseuille flow, with $n = 0.5$.
